---
title: "hw02-MaunishUBC"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Homework 2

```{r warning = FALSE}
library(gapminder)
library(tidyverse)
```

## Smell test the gapminder dataset
```{r}
head(gapminder)
```


What is the type of the gapminder object?
```{r}
typeof(gapminder)
```

What is the class of the gapminder object
```{r}
class(gapminder)
```

How manny rows and columns in the object?

dim
```{r}
dim(gapminder)
```

Rows => Observations:
```{r}
nrow(gapminder)
```

Columns => Variables
```{r}
ncol(gapminder)
```

Using nrow() and ncol() instead of dim() allow to use the individual values in a differnt functions.

What is the data type of each variable

Country is:
```{r}
typeof(gapminder$country)
```

Continent is:
```{r}
typeof(gapminder$continent)
```

Year is:
```{r}
typeof(gapminder$year)
```

LifeExp is:
```{r}
typeof(gapminder$lifeExp)
```

Population is:
```{r}
typeof(gapminder$pop)
```

GDP Per capita is:
```{r}
typeof(gapminder$gdpPercap)
```

## Pick up atleast one catagorical and one quantitative variable to explore:

Catagorical variable: Continent
```{r}
unique(gapminder$continent)
```

Quantitative variable: Population
```{r}

range(gapminder$pop)

```

Spead of Population:
```{r}
ggplot(gapminder, aes(x=pop)) + geom_histogram() +
  scale_x_log10()
```

Summary of the population
```{r}
summary(gapminder$pop)
```

Population on the basis of each continent
```{r}
gapminder %>% 
  ggplot(aes(x=pop)) + geom_histogram() +
  scale_x_log10() + facet_wrap(~ continent)
```
The faceted plot clearty shows that Asia has the highest population

## Exploring various plot types
```{r}
pvsg <- ggplot(gapminder, aes(pop,gdpPercap)) + scale_x_log10() + scale_y_log10()
pvsg + geom_point()
```

```{r}
pvsg + geom_point(aes(colour = continent))
```

This graph shows the population vs gdpPercap and it is intersting to see that Europe and Americas are richer than Africa and Asia.
And in Asia, the recipe for a higher GDP is lower population as we can see for the countries on the top left of the plot.


Modifying the bin size for population
```{r}
ggplot(gapminder, aes(x=pop)) + geom_histogram(bins=50) +
  scale_x_log10() 
```

Making a density plot for the histogram:
```{r}
ggplot(gapminder, aes(x=pop)) + geom_density() +
  scale_x_log10() 
```


Breaking population down on a per continent basis:
```{r}
pvsc <- ggplot(gapminder, aes(continent,pop)) +
  scale_y_log10()
pvsc + geom_boxplot()
```

Let's make a violin plot:
```{r}
pvsc + geom_violin() +
  geom_jitter(alpha = 0.2)
```

```{r}
africa_data <- gapminder %>% filter(continent =='Africa') %>% 
  select(-continent)
```

Plot life exp for each country over time
```{r}
c  <- ggplot(africa_data, aes(year, lifeExp))
c + geom_line()
c + geom_line(aes(group=country))

```

This shows that the life expectancy constantly increases till 1980, after which few countries show a decrease in life exp: 
Especially one that has a life exp below 30!
Which country is it?
```{r}
africa_data %>% filter(lifeExp < 30)
```

So, it's Rwanda and this loss in lifeExp was due to the Rwanda genocide!

Did the genocide impact the population and the gdpPercap?

```{r}
africa_data %>% filter(country == 'Rwanda') %>% 
  ggplot(aes(year, gdpPercap)) + geom_line()
```

So the GDP indeed declining prior to the genocide.

What about the total population? 

```{r}
africa_data %>% filter(country == 'Rwanda') %>% 
  ggplot(aes(year, pop)) + geom_line()
```

The population was rapidly increasing and abruptly platued during the genocide and then started increasing again.
You can read more about the genocide here:

(Vox: Rwanda's genocide — what happened, why it happened, and how it still matters)[https://www.vox.com/2014/4/10/5590646/rwandan-genocide-anniversary]

(BBC: Rwanda genocide: 100 days of slaughter)[https://www.bbc.com/news/world-africa-26875506]

(The New York Times: Rwandan Genocide)[https://www.nytimes.com/topic/destination/rwandan-genocide]