---
title: "hw02-MaunishUBC"
output: github_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## Loading gapinder and tidyverse

```{r warning = FALSE}
library(gapminder)
library(tidyverse)
```

## Smell testing the gapminder dataset
```{r}
head(gapminder)
```


What is the type of the gapminder object?
```{r}
typeof(gapminder)
```

What is the class of the gapminder object?
```{r}
class(gapminder)
```

How many rows and columns are in gapminder?

dim
```{r}
dim(gapminder)
```

Rows => Observations:
```{r}
nrow(gapminder)
```

Columns => Variables
```{r}
ncol(gapminder)
```

Using nrow() and ncol() instead of dim() allows us to use the individual values of in other functions.
eg.you can count from 1:nrow() in an if or while loop.

What is the data type of each variable  in gapminder?

Country is:
```{r}
typeof(gapminder$country)
```

Continent is:
```{r}
typeof(gapminder$continent)
```

Year is:
```{r}
typeof(gapminder$year)
```

LifeExp is:
```{r}
typeof(gapminder$lifeExp)
```

Population is:
```{r}
typeof(gapminder$pop)
```

GDP Per capita is:
```{r}
typeof(gapminder$gdpPercap)
```

## Exploring individual variables

Catagorical variable: Continent
Quantitative variable: Population


What are the continents listed in the gapminder dataset?
```{r}
unique(gapminder$continent)
```

What is the range of population for the countries in the gapminder dataset?
```{r}

range(gapminder$pop)

```

What's the spread of the population in the gapminder dataset?
```{r}
ggplot(gapminder, aes(x=pop)) + geom_histogram() +
  scale_x_log10()
```


As we were recommended to change the bin size for Histogram, let's modify the bin size for population
```{r}
ggplot(gapminder, aes(x=pop)) + geom_histogram(bins=50) +
  scale_x_log10() 
```


Making a density plot for the histogram:
```{r}
ggplot(gapminder, aes(x=pop)) + geom_density() +
  scale_x_log10() 
```

Can you explore more about the nature of the population distribution?
```{r}
summary(gapminder$pop)
```

## Exploring various plot types

```{r}
gapminder %>% 
  ggplot(aes(x=pop)) + geom_histogram(bins=50) +
  scale_x_log10() + facet_wrap(~ continent)
```

The faceted plot clearly shows that Asia has the highest population

How does the gapminder dataset look when you compare the gdpPercap vs population?
```{r}
pvsg <- ggplot(gapminder, aes(pop,gdpPercap)) + scale_x_log10() + scale_y_log10()
pvsg + geom_point()
```

```{r}
pvsg + geom_point(aes(colour = continent))
```

Insights:

- This graph shows the population vs gdpPercap for the gapminder dataset and it is intersting to see that Europe and Americas are richer than Africa and Asia.

- Interstingly in Asia, the recipe for a higher GDP is lower population as we can see for the countries on the top left corner of the plot.


## Plot of one categorical and one quantitative variable

Breaking population down on a per continent basis:
```{r}
pvsc <- ggplot(gapminder, aes(continent,pop)) +
  scale_y_log10()
pvsc + geom_boxplot()
```

Let's make a violin plot and jitter the points:

```{r}
pvsc + geom_violin() +
  geom_jitter(alpha = 0.2)
```


## Use filter, select, pipe and do advanced anaysis: 

```{r warning=FALSE}
africa_data <- gapminder %>% filter(continent =='Africa') %>% 
  select(-continent)
```

Plot life exp for each country over time
```{r}
c  <- ggplot(africa_data, aes(year, lifeExp))
c + geom_line(aes(group=country))

```

This shows that the life expectancy constantly increases till 1980, after which few countries show a decrease in life exp: 
Especially one that has a life exp below 30 after 1990!

Which country is it?
```{r}
africa_data %>% filter(lifeExp < 30)
```

The country is **Rwanda** and what caused this drastic decrease in lifeExp?
Did this drop in lifeExp correspond to fluctuations in population and the gdpPercap?

```{r}
africa_data %>% filter(country == 'Rwanda') %>% 
  ggplot(aes(year, gdpPercap)) + geom_line()
```

So the GDP was declining since the 1980s and reached a local minimum after 1995.

What about the total population? 
```{r}
africa_data %>% filter(country == 'Rwanda') %>% 
  ggplot(aes(year, pop)) + geom_line()
```

The population was rapidly increasing and abruptly platued after 1992 till 1997 then started increasing again.

These drastic changes in lifeExp, gdpPercap and pop were due to the **Rwanda Genocide** in the 1990s where more than a million people died! 

*You can read more about the genocide here:*

1. [Vox: Rwanda's genocide — what happened, why it happened, and how it still matters](https://www.vox.com/2014/4/10/5590646/rwandan-genocide-anniversary)

2. [BBC: Rwanda genocide: 100 days of slaughter](https://www.bbc.com/news/world-africa-26875506)

3. [The New York Times: Rwandan Genocide](https://www.nytimes.com/topic/destination/rwandan-genocide)